Exploring mathematical objects from custom-tailored mathematical universes

### 2020-03-28 paper-filmat.tex
typo

That said, intuitionistic logic still allows for a considerable amount of
freedom, and in many toposes statements are true which are baffling if one has
only received training in mathematics based on classical logic. For instance, on first sight it
looks like the sign function
\[ \operatorname{sgn} : \RR \longrightarrow \RR,\ x \longmapsto \begin{cases}
  -1, & \text{if $x < 0$,} \\
  0, & \text{if $x = 0$,} \\
  1, & \text{if $x > 0$,}
\end{cases} \]
is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
continuous''. However, closer inspection reveals that the sign function
cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
logic is available. The domain of the sign
function is the subset~$\{ x \in \RR \,|\, x < 0 \vee x = 0 \vee x > 0 \}
\subseteq \RR$, and in intuitionistic logic this subset cannot be shown to
coincide with~$\RR$.

@@

What the \effective topos believes to be a real number is, from the external
point of view, a Turing machine~$X$ which outputs, when called with a natural
number~$n$ as input, a rational approximation~$X(n)$. These approximations are
required to be \emph{consistent} in the sense that~$|X(n) - X(m)|
\leq 2^{-n} + 2^{-m}$. Intuitively, such a machine~$X$ denotes the real
number~$\lim_{n \to \infty} X(n)$, and the approximations~$X(n)$ must be
within~$2^{-n}$ of the limit.

@@

The \effective topos inherits Markov's principle from the meta level:
The statement~``$\Eff \models \text{\eqref{eq:markov}}$'' means that there is a
Turing machine~$M$ which, when given the description of a Turing machine~$F$
computing a function~$f : \NN \to \NN$ as input, outputs the description of a Turing
machine~$S_F$ which, when given a witness of~``$\neg\neg \exists n\?\NN\_ f(n)
= 0$'', outputs a witness of~``$\exists n\?\NN\_ f(n) = 0$'' (up to trivial
conversions, a number~$n$ such that~$f(n) = 0$).

@@

\begin{figure}
  \includegraphics[height=3cm]{trichotomy-1}
  \includegraphics[height=3cm]{trichotomy-2}
  \includegraphics[height=3cm]{trichotomy-3}
  \caption{\label{fig:trichotomy} Three examples of what the topos~$\Sh(X)$
  each believes to be a single real number, where the base space~$X$ is the
  unit interval. (a) A positive real number. (b) A negative real number. (c) A
  number which is neither negative nor zero nor positive. Externally speaking,
  there is no covering of the unit interval by open subsets on which the
  depicted function~$a$ is either everywhere negative, everywhere zero or everywhere
  positive.}
\end{figure}

\begin{figure}
  \includegraphics[height=3cm]{zeros-in-families-still}
  \caption{\label{fig:ivt} A single member~$f_{x_0}$ of a continuous
  family~$(f_x)_{x \in X}$ of continuous functions~$f_x : \RR \to \RR$. The
  parameter space is~$X = [0,1]$ (not shown) and the other functions~$f_x$ are
  obtained from~$f_{x_0}$ by moving the horizontal plateau further up or further
  down. The depicted member~$f_{x_0}$ has a unique
  zero and there is an open neighborhood~$U$ of~$x_0$ on which zeros of the
  functions~$f_x$, $x \in U$ can be picked continuously. However, there is no
  such neighborhood of that particular parameter value~$x_1$ for which the
  horizontal plateau lies on the $x$-axis.}
\end{figure}

@@

In contrast, the intermediate value theorem for monotone functions does have an
intuitionistic proof and hence applies in the internal universe of~$\Sh(X)$.
Thus for any continuous family~$(f_x)_x$ of continuous monotone functions
with~$f_x(-1) < 0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible
to pick zeros of the family in a continuous fashion.\footnote{While the
Kripke--Joyal translation of~``$\exists$'' is by definition \emph{local
existence}, one can show that the Kripke--Joyal translation of~``$\exists!$''
is \emph{unique existence on all opens}, in particular \emph{unique global
existence}. Because the conclusion of the intermediate value theorem for
monotone functions can be strengthened from ``has a zero'' to ``has a unique
zero'', this observation shows that the zeros can even globally be picked in a
continuous fashion.}
\end{ex}

@@

\subsection{Hyperreal numbers} There are several ways for introducing
infinitesimals into rigorous mathematics. One is Robinson's \emph{nonstandard
analysis}, where we enlarge the field~$\RR$ of real numbers to a
field~$^\star\RR$ of \emph{hyperreal numbers} by means of a non-principal
ultrafilter.

@@

For reference purposes only, we include the precise definition of the Zariski topos.
\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
the category of functors from finitely presented~$\RR$-algebras to sets, namely of
the Zariski sheaves. Such a functor is a \emph{Zariski sheaf} if and only if,
for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$,
the diagram
\[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
F(A[(f_jf_k)^{-1}]) \]
is a coequalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
\end{defn}

@@

\paragraph{Properties of the smooth numbers}\label{page:field-property}
As a concrete example, the Kripke--Joyal translation of the statement that~$\RR^\sim$ is a field,
\[ \Zar(\RR) \models \forall x\?\RR^\sim\_ (\neg(x = 0) \Rightarrow \exists y\?\RR^\sim\_ xy = 1), \]
is this:
\begin{indentblock}
For any stage~$A$ and any element~$x \in A$, \\
${\qquad}$ for any later stage~$B$ of~$A$, \\
${\qquad\qquad}$ if for any later stage~$C$ of~$B$ \\
${\qquad\qquad\qquad}$ in which~$x = 0$ holds \\
${\qquad\qquad\qquad}$ also~$1 = 0$ holds, \\
${\qquad\qquad}$ then~$B$ can be covered by later stages~$C_i$ such that, \\
${\qquad\qquad\qquad}$ for each
index~$i$, there is an element~$y \in C_i$ with~$xy = 1$ in~$C_i$.
\end{indentblock}
And indeed, this statement is true. Let a stage~$A$ (a finitely
presented~$\RR$-algebra) and an element~$x \in A$ which fulfills the stated
condition be given. Let~$B$ be any later stage of~$A$ (any finitely
presented~$A$-algebra -- such an algebra is also finitely presented as
an~$\RR$-algebra). Then~$x = 0$ holds in the refinement~$C \defeq B/(x)$.
Hence~$1 = 0$ holds in~$C$. By elementary algebra, this means that~$x$ is
invertible in~$B$. Hence the conclusion holds for the singleton covering
of~$B$ by~$C_1 \defeq B$.

>>>>

That said, intuitionistic logic still allows for a considerable amount of
freedom, and in many toposes statements are true which are baffling if one has
only received training in mathematics based on classical logic. For instance, on first sight it
looks like the sign function
\[ \operatorname{sgn} : \RR \longrightarrow \RR,\ x \longmapsto \begin{cases}
  -1, & \text{if $x < 0$,} \\
  0, & \text{if $x = 0$,} \\
  1, & \text{if $x > 0$,}
\end{cases} \]
is an obvious counterexample to the statement ``any map~$\RR \to \RR$ is
continuous''. However, a closer inspection reveals that the sign function
cannot be proven to be a total function~$\RR \to \RR$ if only intuitionistic
logic is available. The domain of the sign
function is the subset~$\{ x \in \RR \,|\, x < 0 \vee x = 0 \vee x > 0 \}
\subseteq \RR$, and in intuitionistic logic this subset cannot be shown to
coincide with~$\RR$.

@@

What the \effective topos believes to be a real number is, from the external
point of view, a Turing machine~$X$ which outputs, when called with a natural
number~$n$ as input, a rational approximation~$X(n)$. These approximations are
required to be \emph{consistent} in the sense that~$|X(n) - X(m)|
\leq 2^{-n} + 2^{-m}$. Intuitively, such a machine~$X$ denotes the real
number~$\lim_{n \to \infty} X(n)$, and each approximation~$X(n)$ must be
within~$2^{-n}$ of the limit.

@@

The \effective topos inherits Markov's principle from the meta level:
The statement~``$\Eff \models \text{\eqref{eq:markov}}$'' means that there is a
Turing machine~$M$ which, when given the description of a Turing machine~$F$
computing a function~$f : \NN \to \NN$ as input, outputs the description of a Turing
machine~$S_F$ which, when given a witness of~``$\neg\neg \exists n\?\NN\_ f(n)
= 0$'', outputs a witness of~``$\exists n\?\NN\_ f(n) = 0$'' (up to trivial
conversions, this is a number~$n$ such that~$f(n) = 0$).

@@

\begin{figure}
  \includegraphics[height=3cm]{trichotomy-1}
  \includegraphics[height=3cm]{trichotomy-2}
  \includegraphics[height=3cm]{trichotomy-3}
  \caption{\label{fig:trichotomy} Three examples of what the topos~$\Sh(X)$
  each believes to be a single real number, where the base space~$X$ is the
  unit interval. (a) A positive real number. (b) A negative real number. (c) A
  number which is neither negative nor zero nor positive. Externally speaking,
  there is no covering of the unit interval by opens on which the
  depicted function~$a$ is either everywhere negative, everywhere zero or everywhere
  positive.}
\end{figure}

\begin{figure}
  \includegraphics[height=3cm]{zeros-in-families-still}
  \caption{\label{fig:ivt} A single member~$f_{x_0}$ of a continuous
  family~$(f_x)_{x \in X}$ of continuous functions~$f_x : \RR \to \RR$. The
  parameter space is~$X = [0,1]$ (not shown) and the other functions~$f_x$ are
  obtained from~$f_{x_0}$ by moving the horizontal plateau further up or
  down. The depicted member~$f_{x_0}$ has a unique
  zero and there is an open neighborhood~$U$ of~$x_0$ on which zeros of the
  functions~$f_x$, $x \in U$ can be picked continuously. However, there is no
  such neighborhood of that particular parameter value~$x_1$ for which the
  horizontal plateau lies on the $x$-axis.}
\end{figure}

@@

In contrast, the intermediate value theorem for (strictly) monotone functions does have an
intuitionistic proof and hence applies in the internal universe of~$\Sh(X)$.
Thus for any continuous family~$(f_x)_x$ of continuous monotone functions
with~$f_x(-1) < 0$ and~$f_x(1) > 0$ for all~$x \in X$, it is locally possible
to pick zeros of the family in a continuous fashion.\footnote{While the
Kripke--Joyal translation of~``$\exists$'' is by definition \emph{local
existence}, one can show that the Kripke--Joyal translation of~``$\exists!$''
is \emph{unique existence on all opens}, in particular \emph{unique global
existence}. Because the conclusion of the intermediate value theorem for
monotone functions can be strengthened from ``has a zero'' to ``has a unique
zero'', this observation shows that the zeros can even globally be picked in a
continuous fashion.}
\end{ex}

@@

\subsection{Hyperreal numbers} There are several ways of introducing
infinitesimals into rigorous mathematics. One is Robinson's \emph{nonstandard
analysis}, where we enlarge the field~$\RR$ of real numbers to a
field~$^\star\RR$ of \emph{hyperreal numbers} by means of a non-principal
ultrafilter.

@@

For reference purposes only, we include the precise definition of the Zariski topos.
\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is a certain full subcategory of
the category of functors from finitely presented~$\RR$-algebras to sets, namely of
the Zariski sheaves. Such a functor is a \emph{Zariski sheaf} if and only if,
for any covering~$(A[f_i^{-1}])_i$ of any finitely presented~$\RR$-algebra~$A$
(this notion is defined in Table~\ref{table:zariski}), the diagram
\[ F(A) \longrightarrow \prod_i F(A[f_i^{-1}]) \xbigtoto{} \prod_{j,k}
F(A[(f_jf_k)^{-1}]) \]
is a coequalizer diagram. The object~$\RR^\sim$ of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
\end{defn}

@@

\paragraph{Properties of the smooth numbers}\label{page:field-property}
As a concrete example, the Kripke--Joyal translation of the statement that~$\RR^\sim$ is a field,
\[ \Zar(\RR) \models \forall x\?\RR^\sim\_ (\neg(x = 0) \Rightarrow \exists y\?\RR^\sim\_ xy = 1), \]
is this:
\begin{indentblock}
For any stage~$A$ and any element~$x \in A$, \\
${\qquad}$ for any later stage~$B$ of~$A$, \\
${\qquad\qquad}$ if for any later stage~$C$ of~$B$ \\
${\qquad\qquad\qquad}$ in which~$x = 0$ holds \\
${\qquad\qquad\qquad}$ also~$1 = 0$ holds, \\
${\qquad\qquad}$ then~$B$ can be covered by later stages~$C_i$ such that, \\
${\qquad\qquad\qquad}$ for each
index~$i$, there is an element~$y \in C_i$ with~$xy = 1$ in~$C_i$.
\end{indentblock}
And indeed, this statement is true. Let a stage~$A$ (a finitely
presented~$\RR$-algebra) and an element~$x \in A$
be given. Let~$B$ be any later stage of~$A$ (any finitely
presented~$A$-algebra -- such an algebra is also finitely presented as
an~$\RR$-algebra). Assume that for any later stage~$C$ of~$B$ in which~$x = 0$
holds also~$1 = 0$ holds. Trivially,~$x = 0$ holds in the particular refinement~$C \defeq B/(x)$.
Hence~$1 = 0$ holds in~$C$. By elementary algebra, this means that~$x$ is
invertible in~$B$. Hence the conclusion holds for the singleton covering
of~$B$ given by~$C_1 \defeq B$.
